---
title: "Untitled"
author: "Trent Meyer"
date: "2022-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)

wc_forecasts <- read_csv("data/world-cup-2018/wc_forecasts.csv")
wc_matches <- read_csv("data/world-cup-2018/wc_matches.csv")
```


Given each team’s SPI rating, the process for generating win/loss/draw probabilities for a World Cup match is three-fold:

1. We calculate the number of goals that we expect each team to score during the match. These projected match scores represent the number of goals that each team would need to score to keep its offensive rating exactly the same as it was going into the match.


2. Using our projected match scores and the assumption that goal scoring in soccer follows a Poisson process, which is essentially a way to model random events at a known rate, we generate two Poisson distributions around those scores. Those give us the likelihood that each team will score no goals, one goal, two goals, etc.

# Russia
```{r}
games <- 0:5
plot(games, dpois(games, 2.03), type = "h")
russiawin <- dpois(games, 2.03)
```

# Saudi Arabia
```{r}
plot(games, dpois(games, 0.73), type = "h")
saudiwin <- dpois(games, 0.73)
```

3. We take the two Poisson distributions and turn them into a matrix of all possible match scores from which we can calculate the likelihood of a win, loss or draw for each team. To avoid undercounting draws, we increase the corresponding probabilities in the matrix.


```{r}
# predicting the probabilities that Russia will score 0:5 goals
russiawin <- dpois(0:5, wc_matches$proj_score1[1])

# predicting the probabilities that Saudi Arabia will score 0:5 goals
saudiwin <- dpois(0:5, wc_matches$proj_score2[1])
```

```{r}
# Turning the probabilities into matrices
russiamatrix <- matrix(russiawin, nrow = 1, ncol = 6)
saudimatrix <- matrix(saudiwin, nrow = 6, ncol = 1)

# Multiplying the matrices to get a 6x6 matrix with probabilities
# that each team will score that amount of goals, from which 
# we can get the probability that each team will win
probmatrix <- saudimatrix %*% russiamatrix 

# I transposed this because using as.vector went down the columns, except I wanted
# to go across the rows first
probmatrix_t <- t(probmatrix)

# With the transposed matrix, I wanted to create a vector, and then use cumsum()
# to get the cumulative sum
probvector <- as.vector(probmatrix, mode = "numeric")
cumprob <- cumsum(probvector)
cumprobmatrix <- matrix(cumprob, nrow = 6, ncol = 6, byrow = TRUE)

rowindices <- row(cumprobmatrix)
rowvector <- as.vector(rowindices, mode = "numeric")

colindices <- col(cumprobmatrix)
colvector <- as.vector(colindices, mode = "numeric")

probindextib <- tibble(rowvector, colvector, cumprob)

# generate a random uniform value
prob_index <- runif(1, 0, 1)

# find out which values fall above that probability
which_index <- which(cumprobmatrix >= prob_index)

# single out the first index, and determine if that would be a win, loss, or draw,
# and for which team
index_sim <- which_index[1]

probindextib %>% slice(index_sim)

lower.tri(cumprobmatrix, diag = FALSE)
upper.tri(cumprobmatrix, diag = FALSE)
diag(cumprobmatrix)



# team1_index_win <- c(2:6, 9:12, 16:18, 23:24, 30)
# index_draw <- c(1, 8, 15, 22, 29, 36)
# team2_index_win <- c(7, 13:14, 19:21, 25:28, 31:35)

# create team1point and team2points x
# pull()
# mutate first
# iterate one matchup, and then iterate for all of the teams
# send an email when I push, figure out iterating for one matchup for Tuesday
# make a function where you return a single value where the team
# replicate something in purrr
# modern data science with R 7.3 map()

probindextib <- probindextib %>% mutate(result = case_when(
  rowvector > colvector ~ "Team 1 Wins",
  rowvector == colvector ~ "Draw",
  rowvector < colvector ~ "Team 2 Wins"
))

probindextib$result[index_sim]

probindextib <- probindextib %>% 
  mutate(team1points = case_when(result == "Draw" ~ 1,
                                                result == "Team 1 Wins" ~ 3,
                                                result == "Team 2 Wins" ~ 0)) %>%
  mutate(team2points = case_when(result == "Draw" ~ 1,
                                                result == "Team 2 Wins" ~ 3,
                                                result == "Team 1 Wins" ~ 0))
```


```{r}
proj_score1 <- 4
proj_score2 <- 0.1
predict_team1 <- function(proj_score1, proj_score2) {
  goals <- 0:10
  # predicting the probabilities that team1 will score 0:10 goals
  team1goals <- dpois(goals, proj_score1)

  # predicting the probabilities that team2 will score 0:10 goals
  team2goals <- dpois(goals, proj_score2)
  
  # Turning the probabilities into matrices
  team1goals_matrix <- matrix(team1goals, nrow = 1, ncol = 11)
  team2goals_matrix <- matrix(team2goals, nrow = 11, ncol = 1)
  
  # Multiplying the matrices to get an 11x11 matrix with probabilities
  # that each team will score that amount of goals, from which 
  # we can get the probability that each team will win
  goalsprobmatrix <- team2goals_matrix %*% team1goals_matrix
  
  # I transposed this because using as.vector went down the columns, except I wanted
  # to go across the rows first
  #goalsprobmatrix_t <- t(goalsprobmatrix)
  
  # With the transposed matrix, I wanted to create a vector, and then use cumsum()
  # to get the cumulative sum
  goalsprobvector <- as.vector(t(goalsprobmatrix), mode = "numeric")
  cumprobmatrix <- matrix(cumsum(goalsprobvector), nrow = 11, ncol = 11, byrow = TRUE)
  cumprob <- as.vector(t(cumprobmatrix), mode = "numeric")
  
  rowindices <- row(cumprobmatrix)
  rowvector <- as.vector(rowindices, mode = "numeric")

  colindices <- col(cumprobmatrix)
  colvector <- as.vector(colindices, mode = "numeric")

  probindextib <- tibble(rowvector, colvector, cumprob)

  # generate a random uniform value
  prob_index <- runif(1, 0, 1)
  
  # find out which values fall above that probability
  which_index <- which(cumprob >= prob_index)

  # single out the first index, and determine if that would be a win, loss, or draw,
  # and for which team
  index_sim <- which_index[1]
  
  probindextib <- probindextib %>% mutate(
    result = case_when(
      rowvector > colvector ~ "Team 1 Wins",
      rowvector == colvector ~ "Draw",
      rowvector < colvector ~ "Team 2 Wins"
    ),
    team1points = case_when(
      result == "Draw" ~ 1,
      result == "Team 1 Wins" ~ 3,
      result == "Team 2 Wins" ~ 0
    ),
    team2points = case_when(
      result == "Draw" ~ 1,
      result == "Team 2 Wins" ~ 3,
      result == "Team 1 Wins" ~ 0
    )
  ) 
  
  result_list <- c(probindextib$team1points[index_sim])
  
  # add in summing up the number of points for each team
  
  return(result_list)
}
```

```{r}
predict_team1(proj_score1 = 2, proj_score2 = 2)

## one thing at a time:
## predict the points for team 1 just for the fifth row of the data set one time
wc_iter <- wc_matches |> select(proj_score1, proj_score2, everything()) |> slice(1, 2, 17, 19, 33, 34)
#wc_iter <- wc_matches |> select(proj_score1, proj_score2, everything()) |> slice(1:48)


predict_team1(proj_score1 = wc_iter |> slice(5) |> pull(1),
              proj_score2 = wc_iter |> slice(5) |> pull(2))


## expand slowly: predict the score for team 1 for the fifth row in the data set a thousand times
niter <- 1000
oneteam_res <- map(1:niter, ~ predict_team1(proj_score1 = wc_iter |> slice(5) |> pull(1),
              proj_score2 = wc_iter |> slice(5) |> pull(2))) |>
  unlist()


## expand slowly: predict the score for team 1 for every row of the data set:
## wrap in for loop and change 5 to i
niter <- 100 ## so messing around with code takes less time
result_matrix <- matrix(NA, nrow = nrow(wc_iter), ncol = niter)

for (i in 1:nrow(wc_iter)) {
  result_matrix[i, ] <- map(1:niter, ~ predict_team1(proj_score1 = wc_iter |> slice(i) |> pull(1),
              proj_score2 = wc_iter |> slice(i) |> pull(2))) |>
  unlist()
}

result_matrix
## each row is a match
## each column is the number of points the first team earned in that iteration

## can keep expanding __slowly__: focus on figuring out who is moving on in each iteration for _1_
## group (6 matches). Once that is working, expand to figuring out who is moving on in each group

## this is getting the average team1points for the each match

## Russia v Saudi Arabia
match1_team1_result <- result_matrix[1,]
sum(match1_team1_result)/100

## Egypt v Uruguay
match2_team1_result <- result_matrix[2,]
sum(match2_team1_result)/100

## Russia v Egypt
match3_team1_result <- result_matrix[3,]
sum(match3_team1_result)/100

## Uruguay v Saudi Arabia
match4_team1_result <- result_matrix[4,]
sum(match4_team1_result)/100

## Saudi Arabia v Egypt
match5_team1_result <- result_matrix[5,]
sum(match5_team1_result)/100

## Uruguay v Russia
match6_team1_result <- result_matrix[6,]
sum(match6_team1_result)/100

## Russia: 2.07, 2.05
## Saudi Arabia: 0.8, 0.82
## Egypt: was not team1
## Uruguay: 2.56, 1.35

## How to figure out what Egypt got, and where to go from here
## Each team should have number of points for three games, however not each team
## was team1 every match they were in.
```



```{r}
#prediction(proj_score1 = 2.03, proj_score2 = 0.73) :: map()
#times <- 1:10
#match1pred <- purrr::map(times, prediction(proj_score1 = 2.03, proj_score2 = 0.73))
##purrr::map(1:10, ~ prediction(proj_score1 = 2.03, proj_score2 = 0.73))
proj_score1 = 1; proj_score2 = 2
match1rep <- tibble(outcome = replicate(n = 1000, prediction(proj_score1, proj_score2)))
match1rep <- match1rep %>% group_by(outcome) %>% 
  summarise(total = n()) %>% 
  mutate(prop = total / 1000)


wc_iter <- wc_matches |> select(proj_score1, proj_score2, everything()) |> slice(1:48)
wc_iter

niter <- 10

results <- rep(NA, nrow(wc_iter))





proj_score1list <- list(wc_matches$proj_score1)

for (i in 1:2) {
  results_vec <-
    purrr::map(1:niter,
               ~ prediction(
                 proj_score1 = wc_iter[i,] |> pull(proj_score1),
                 proj_score2 = wc_iter[i,] |> pull(proj_score2)
               )) |>
    unlist()
  
  results_mat[i] <-
    tibble(outcome = results_vec) |> group_by(outcome) |>
    summarise(total = n()) |>
    mutate(prop = total / niter)
}


results_mat[1]
map(1:1000, ~ prediction(proj_score1 = 2.03, proj_score2 = 0.73)) |>
  unlist()
```

