---
title: "World Cup Prediction Report"
author: "Trent Meyer"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# World Cup Prediction
## Introduction

|             The FIFA World Cup is an event that grips the attention of soccer (or fútbol, football depending on where you are in the world) fans all across the world. This event occurs once every four years, bringing together millions of individuals from six different continents and a wide range of cultures. To have your nation qualify for the World Cup is quite prestigious as only 32 countries (48 in 2026) are able to do so. This rigorous qualifying schedule involves playing the top teams in your region over the course of about two years prior to the next World Cup. 

|             The United States, despite not being completely enthralled like say Brazil, still nonetheless sees a magic atmosphere surrounding the tournament. However, in 2018 the United States lost to Trinidad & Tobago, a team they were supposed to easily beat, in their last qualifying game which effectively knocked them out of that cycle's tournament. This sparked quite a long chain of events in the United States Soccer Federation (USSF, the main governing body for soccer and the national teams) over the next four years. This included completely revamping the youth development system, hiring a new manager, and phasing out all of the aging players in favor for bright, up and coming stars making their presence known in European academies.

|             This chain of events changed quite a lot for the USSF, as seen since 2018 and today. Our men's national team saw its stars playing at some of the most highly regarded clubs in the world such as Juventus, Chelsea, and Borussia Dortmund. Not only to mention the quality of the clubs, these players were teenagers starting and making an impact for their club. We saw a higher number of teenagers playing in Europe than ever before, meaning our players were developing at the best academies in Europe. Over the next few years, our young team managed to win two regional tournaments and secure our spot in the 2022 World Cup in Qatar. 

|             With such a large event, many want to predict the winner of this prestigious trophy. After 22 tournaments, only eight have managed to win the tournament. Many companies such as FiveThirtyEight and ESPN, to name only a couple, attempt to predict the winner of the upcoming tournament using many different methods. 

|             [FiveThirtyEight](https://fivethirtyeight.com/features/how-our-2018-world-cup-predictions-work/) created a process to predict the winner of the 2018 tournament. This process is explained, along with the data being available in a [GitHub repository](https://github.com/fivethirtyeight/data/tree/master/world-cup-2018). Using their data, I am attempting to recreate the probabilities that each team will win each game in the group stage.

## Methods

|             To begin their predictions, each team's World Cup roster was given an SPI rating, which judged the strength of that team. Each team was given an offensive rating, which is "the number of goals that it would be expected to score against an average team on a neutral field" (Boice, 2018). The defensive rating followed the same situation, except the number of goals they would be expected to concede against an average team. The overall "match" SPI rating is the "percentage of points ... the team would be expected to take if the match were played over and over again" (Boice, 2018). This SPI rating is calculated using a database of international matches dating back to 1905! They also calculated a roster-based SPI rating which looked at each player, and how much they played, in what league, and how elite the competition they are facing each week is. 

|             Once the SPI rating was calculated, they began by predicting the match scores, which are the number of goals that team would need to score to uphold their rating. Using the projected score, they completed a Poisson process where they predicted the probabilities that each team would score one goal, two goals, and so on up to ten goals (a little uncommon, but not outlandish). 

|             The next step was to convert these to matrices, and multiply them together to determine the likelihood that one team would score x amount of goals *and* the other team would score y amount of goals. Using this matrix, I was able to calculate the cumulative probability, and generate a random value that would correspond to an outcome based on the matrix. Each match in the group stages was iterated 1000 times, from which we could determine the number of points each team would score in that iteration. Next, I was able to calculate the proportion of times that each team would advance out of the group stages (place in the top two out of four teams in their group).

## Results

```{r, fig.cap="Figure 1: Proportion of times each team advanced out of their respective group"}
library(tidyverse)
load("wc_long_iter_total.rda")
ggplot(data = wc_long_iter_total, aes(x = proptimes, y = fct_reorder(team, proptimes))) +
  geom_point() + 
  geom_segment(data = wc_long_iter_total, aes(x = 0, xend = proptimes, y = team, yend = team)) +
  facet_wrap(~group, scales = "free_y")
```

|             From Figure 1, we are able to see that the teams that made it through the highest proportion of times were Argentina, Brazil, England as well as a few others. All of these teams have previously won the World Cup, and made it out of their groups in the 2018 tournament. The teams that were predicted to make it out the least number of times were Australia, Iran, and South Korea to name a few.

## Discussion

|             One of the more difficult aspects of this was that I was not able to completely replicate the probability matrix, as their article only gave limited information on the process. However, with the information given, I learned a lot about predicting World Cup games using skills I learned in linear algebra, probability, and data science. I had to begin by predicting only one game, and once that process was completed, I was able to expand out to one group of four teams. 

|             This process took much longer than I expected, and given I only had a semester to work, I was only able to predict the group stages. If given more time, I would be able to figure out a way to set up the knockout stage bracket, and then use the expected scores given to predict who would win that World Cup.

## Appendix